---
title: "Assignment 5: Data Visualization"
author: "GuruBandaa Khalsa"
output: pdf_document
geometry: margin=2.54cm
---

```{r setup, include=FALSE}
#This code chunk will tidy your knit PDF files, wrapping long code lines
#For it to work, the "formatR" package needs to be installed

#install.packages('formatR')
knitr::opts_chunk$set(tidy.opts=list(width.cutoff=45), tidy=TRUE)
```

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on Data Visualization 

## Directions
1. Rename this file `<FirstLast>_A05_DataVisualization.Rmd` (replacing `<FirstLast>` with your first and last name).
2. Change "Student Name" on line 3 (above) with your name.
3. Work through the steps, **creating code and output** that fulfill each instruction.
4. Be sure to **answer the questions** in this assignment document.
5. When you have completed the assignment, **Knit** the text and code into a single PDF file.

The completed exercise is due on Friday, Oct 21st @ 5:00pm.



## Set up your session 
1. Set up your session. Verify your working directory and load the tidyverse, 
lubridate, & cowplot packages. Upload the NTL-LTER processed data files for  
nutrients and chemistry/physics for Peter and Paul Lakes (use the tidy 
[`NTL-LTER_Lake_Chemistry_Nutrients_PeterPaul_Processed.csv`] version) and 
the processed data file for the Niwot Ridge litter dataset (use the
[`NEON_NIWO_Litter_mass_trap_Processed.csv`] version). 

2. Make sure R is reading dates as date format; if not change the format to date.

```{r}
#1. I checked my working directory using the getwd() function, loaded the
# `tidyverse`, `lubridate`, and cowplot packages, and uploaded the NTL-LTER
# processed data files for nutrients and chemistry/physics for Peter and Paul Lakes.
getwd()
library(tidyverse)
library(lubridate)
library(cowplot)
NTL1 <- read.csv("~/Desktop/Masters at Duke/Second Year/Fall Semester/Environmental Data Analytics/EDA-Fall2022/Data/Processed/NTL-LTER_Lake_Chemistry_Nutrients_PeterPaul_Processed.csv", stringsAsFactors=TRUE)
NTL2 <- read.csv("~/Desktop/Masters at Duke/Second Year/Fall Semester/Environmental Data Analytics/EDA-Fall2022/Data/Processed/NEON_NIWO_Litter_mass_trap_Processed.csv", stringsAsFactors=TRUE)

#2. I used the View() function to confirm that R is reading dates as date format.
View(NTL1)
View(NTL2)
```
## Define your theme
3. Build a theme and set it as your default theme.

```{r}
#3. I assigned a pre-built theme from Ggplot as a common theme across all plots
# in my analysis session.
mytheme <- theme_classic(base_size = 14) +
  theme(axis.text = element_text(color = "black"), 
        legend.position = "top")
```

## Create graphs
For numbers 4-7, create ggplot graphs and adjust aesthetics to follow best 
practices for data visualization. Ensure your theme, color palettes, axes, 
and additional aesthetics are edited accordingly.

4. [NTL-LTER] Plot total phosphorus (`tp_ug`) by phosphate (`po4`), with 
separate aesthetics for Peter and Paul lakes. Add a line of best fit and 
color it black. Adjust your axes to hide extreme values (hint: change the 
limits using `xlim()` and/or `ylim()`).

```{r}
#4. I created a ggplot graph with appropriate adjusted aesthetics for total
# phosphorus (`tp_ug`) by phosphate (`po4`), with separate aesthetics for Peter
# and Paul lakes.

```


5. [NTL-LTER] Make three separate boxplots of (a) temperature, (b) TP, and 
(c) TN, with month as the x axis and lake as a color aesthetic. Then, create 
a cowplot that combines the three graphs. Make sure that only one legend is 
present and that graph axes are aligned. 

Tip: R has a build in variable called `month.abb` that returns a list of months;
see <https://r-lang.com/month-abb-in-r-with-example>

```{r}
#5. I created three separate boxplots of (a) temperature, (b) TP, and 
# (c) TN, with month as the x axis and lake as a color aesthetic.  I also
# created a cowplot that combines the three graphs.

```

Question: What do you observe about the variables of interest over seasons 
and between lakes? 

> Answer: 

6. [Niwot Ridge] Plot a subset of the litter dataset by displaying only the 
"Needles" functional group. Plot the dry mass of needle litter by date and 
separate by NLCD class with a color aesthetic. (no need to adjust the name 
of each land use)

7. [Niwot Ridge] Now, plot the same plot but with NLCD classes separated into 
three facets rather than separated by color. 

```{r}
#6. I plotted a subset of the litter dataset by displaying only the "Needles"
# functional group.  I also plotted the dry mass of needle litter by date and 
# separated by NLCD class with a color aesthetic.


#7. I created the same plot as number 6. but with NLCD classes separated into 
# three facets rather than separated by color.


```
Question: Which of these plots (6 vs. 7) do you think is more effective, and why? 

> Answer: