---
title: "Assignment 2: Coding Basics"
author: "GuruBandaa Khalsa"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on coding basics.

## Directions

1.  Rename this file `<FirstLast>_A02_CodingBasics.Rmd` (replacing `<FirstLast>` with your first and last name).
2.  Change "Student Name" on line 3 (above) with your name.
3.  Work through the steps, **creating code and output** that fulfill each instruction.
4.  Be sure to **answer the questions** in this assignment document.
5.  When you have completed the assignment, **Knit** the text and code into a single PDF file.
6.  After Knitting, submit the completed exercise (PDF file) to Sakai.

## Basics Day 1

1.  Generate a sequence of numbers from one to 100, increasing by fours. Assign this sequence a name.

2.  Compute the mean and median of this sequence.

3.  Ask R to determine whether the mean is greater than the median.

4.  Insert comments in your code to describe what you are doing.

```{r}
#1. I used the sequence function to generate a sequence of numbers from one to 100, increasing by fours.  I also assigned the sequence the name "four_sequence."
seq(1, 100, 4)
four_sequence <- seq(1, 100, 4)

#2. I used the mean and median functions to compute the mean and median of the sequence created above, "four_sequence."  I then assigned the mean function the name "seqmean" and the median sequence the name "seqmedian" so that they would be easy to call in the following conditional statement.
mean(four_sequence)
median(four_sequence)
seqmean <- mean(four_sequence)
seqmedian <- median(four_sequence)

#3. I used a conditional statement to ask R to determine whether the mean is greater than the median or if they are equal to one another.
if (seqmean > seqmedian) {
  print("The mean is greater than the median.")
} else if (seqmean < seqmedian) {
  print("The median is greater than the mean.")
} else {
  print("The mean and median are equal.")
}

#4. I inserted comments in my code to describe what I am doing.
```

## Basics Day 2

5.  Create a series of vectors, each with four components, consisting of (a) names of students, (b) test scores out of a total 100 points, and (c) whether or not they have passed the test (TRUE or FALSE) with a passing grade of 50.

6.  Label each vector with a comment on what type of vector it is.

7.  Combine each of the vectors into a data frame. Assign the data frame an informative name.

8.  Label the columns of your data frame with informative titles.

```{r}
#5. I created three vectors, each with four components, titled "students," "scores," and "passing."
students <- c("Jimmy", "Johnny", "Timmy", "Victor")  # The vector "students" is a character vector.
scores <- c(99, 76, 45, 95)  # The vector "scores" is a double vector.
passing <- c(TRUE, TRUE, FALSE, TRUE)  # The vector "passing" is a logical vector.

#6. I used the typeof function to determine the type of each vector and labeled each vector with a comment on what type of vector it is below and in the step above.
typeof(students)  # The vector "students" is a character vector.
typeof(scores)  # The vector "scores" is a double vector.
typeof(passing)  # The vector "passing" is a logical vector.

#7. I combined each of the vectors into a data frame and assigned the data frame the informative name "".  I then viewed the data frame.
test_results <- data.frame(students, scores, passing)
test_results

#8. I labeled the columns of my data frame with the informative titles: "Students," "Scores," and "Passing."  I then viewed the data frame again.
colnames(test_results) <- c("Students", "Scores", "Passing")
test_results
```

9.  QUESTION: How is this data frame different from a matrix?

> Answer: This data frame is different from a matrix because this data frame contains three classes of data (character, double, and logical vectors).  Matrices can only manage a single class of data.

10. Create a function with an if/else statement. Your function should take a **vector** of test scores and print (not return) whether a given test score is a passing grade of 50 or above (TRUE or FALSE). You will need to choose either the `if` and `else` statements or the `ifelse` statement. 

11. Apply your function to the vector with test scores that you created in number 5.

```{r}
#10. I created a function with an if/else statement that takes a vector of test scores and prints whether a given test score is a passing grade of 50 or above.
tester <- function(x) {
  tester2 <- ifelse(x > 50, "TRUE", "FALSE")
  print(tester2)
}

#11. I applied my function to the vector with test scores that I created in number 5.
tester(scores)
```

12. QUESTION: Which option of `if` and `else` vs. `ifelse` worked? Why?

> Answer: "If" and "else" did not work and "ifelse" did work.  This is because ifelse statements are a vector equivalent form of the "if" "else" statement.  This was necessary in this scenario since we are working with a vector.
