---
title: "Assignment 4: Data Wrangling"
author: "GuruBandaa Khalsa"
output: pdf_document
geometry: margin=2.54cm
---

```{r setup, include=FALSE}
#This code chunk will tidy your knit PDF files, wrapping long code lines
#For it to work, the "formatR" package needs to be installed

#install.packages('formatR')
knitr::opts_chunk$set(tidy.opts=list(width.cutoff=80), tidy=TRUE)
```

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on Data Wrangling 

## Directions
1.  Rename this file `<FirstLast>_A03_DataExploration.Rmd` (replacing `<FirstLast>` with your first and last name).
2. Change "Student Name" on line 3 (above) with your name.
3. Work through the steps, **creating code and output** that fulfill each instruction.
4. Be sure to **answer the questions** in this assignment document.
5. When you have completed the assignment, **Knit** the text and code into a single PDF file.

The completed exercise is due on Friday, Oct7th @ 5:00pm.

## Set up your session 
1. Check your working directory, load the `tidyverse` and `lubridate` packages, and upload all four raw data files associated with the EPA Air dataset, being sure to set string columns to be read in as factors. See the README file for the EPA air datasets for more information (especially if you have not worked with air quality data previously). 

2. Explore the dimensions, column names, and structure of the datasets. 

```{r, message = FALSE}
#1. I checked my working directory using the getwd() function, loaded the `tidyverse` and `lubridate` packages, and uploaded all four raw data files associated with the EPA Air dataset.
getwd()
library(tidyverse)
library(lubridate)
EPA1 <- read.csv("./Data/Raw/EPAair_O3_NC2018_raw.csv", stringsAsFactors=TRUE)
EPA2 <- read.csv("./Data/Raw/EPAair_O3_NC2019_raw.csv", stringsAsFactors=TRUE)
EPA3 <- read.csv("./Data/Raw/EPAair_PM25_NC2018_raw.csv", stringsAsFactors=TRUE)
EPA4 <- read.csv("./Data/Raw/EPAair_PM25_NC2019_raw.csv", stringsAsFactors=TRUE)

#2. I explored the dimensions, column names, and structure of the datasets using the dim(), View(), and str() functions.
dim(EPA1)
dim(EPA2)
dim(EPA3)
dim(EPA4)

colnames(EPA1)
colnames(EPA2)
colnames(EPA3)
colnames(EPA4)

str(EPA1)
str(EPA2)
str(EPA3)
str(EPA4)
```

## Wrangle individual datasets to create processed files.
3. Change date to date
4. Select the following columns: Date, DAILY_AQI_VALUE, Site.Name, AQS_PARAMETER_DESC, COUNTY, SITE_LATITUDE, SITE_LONGITUDE
5. For the PM2.5 datasets, fill all cells in AQS_PARAMETER_DESC with "PM2.5" (all cells in this column should be identical).
6. Save all four processed datasets in the Processed folder. Use the same file names as the raw files but replace "raw" with "processed".

```{r}
#3. I changed date to date.
EPA1$Date <- as.Date(EPA1$Date, format = "%m/%d/%Y")
EPA2$Date <- as.Date(EPA2$Date, format = "%m/%d/%Y")
EPA3$Date <- as.Date(EPA3$Date, format = "%m/%d/%Y")
EPA4$Date <- as.Date(EPA4$Date, format = "%m/%d/%Y")

#4. I used the select() function to select the following columns: Date, DAILY_AQI_VALUE, Site.Name, AQS_PARAMETER_DESC, COUNTY, SITE_LATITUDE, SITE_LONGITUDE
EPA1.select <- select(EPA1, Date, DAILY_AQI_VALUE, Site.Name, AQS_PARAMETER_DESC, COUNTY, SITE_LATITUDE, SITE_LONGITUDE)
EPA2.select <- select(EPA2, Date, DAILY_AQI_VALUE, Site.Name, AQS_PARAMETER_DESC, COUNTY, SITE_LATITUDE, SITE_LONGITUDE)
EPA3.select <- select(EPA3, Date, DAILY_AQI_VALUE, Site.Name, AQS_PARAMETER_DESC, COUNTY, SITE_LATITUDE, SITE_LONGITUDE)
EPA4.select <- select(EPA4, Date, DAILY_AQI_VALUE, Site.Name, AQS_PARAMETER_DESC, COUNTY, SITE_LATITUDE, SITE_LONGITUDE)

#5. I filled all cells in AQS_PARAMETER_DESC with "PM2.5" in the PM2.5 datasets.
EPA3.select$AQS_PARAMETER_DESC <- "PM2.5"
EPA4.select$AQS_PARAMETER_DESC <- "PM2.5"

#6. I renamed all four processed datasets and used the write.csv() function to save them in the Processed folder.
EPAair_O3_NC2018_processed.csv <- EPA1.select
EPAair_O3_NC2019_processed.csv <- EPA2.select
EPAair_PM25_NC2018_processed.csv <- EPA3.select
EPAair_PM25_NC2019_processed.csv <- EPA4.select

write.csv(EPAair_O3_NC2018_processed.csv, row.names = FALSE, file = "./Data/Processed/EPAair_O3_NC2018_processed.csv")
write.csv(EPAair_O3_NC2019_processed.csv, row.names = FALSE, file = "./Data/Processed/EPAair_O3_NC2019_processed.csv")
write.csv(EPAair_PM25_NC2018_processed.csv, row.names = FALSE, file = "./Data/Processed/EPAair_PM25_NC2018_processed.csv")
write.csv(EPAair_PM25_NC2019_processed.csv, row.names = FALSE, file = "./Data/Processed/EPAair_PM25_NC2019_processed.csv")
```


## Combine datasets
7. Combine the four datasets with `rbind`. Make sure your column names are identical prior to running this code. 
8. Wrangle your new dataset with a pipe function (%>%) so that it fills the following conditions: 

* Include all sites that the four data frames have in common: "Linville Falls", "Durham Armory", "Leggett", "Hattie Avenue", "Clemmons Middle", "Mendenhall School", "Frying Pan Mountain", "West Johnston Co.", "Garinger High School", "Castle Hayne", "Pitt Agri. Center", "Bryson City", "Millbrook School" (the function `intersect` can figure out common factor levels)
* Some sites have multiple measurements per day. Use the split-apply-combine strategy to generate daily means: group by date, site, aqs parameter, and county. Take the mean of the AQI value, latitude, and longitude.
* Add columns for "Month" and "Year" by parsing your "Date" column (hint: `lubridate` package)
* Hint: the dimensions of this dataset should be 14,752 x 9.

9. Spread your datasets such that AQI values for ozone and PM2.5 are in separate columns. Each location on a specific date should now occupy only one row.
10. Call up the dimensions of your new tidy dataset.
11. Save your processed dataset with the following file name: "EPAair_O3_PM25_NC1718_Processed.csv"

```{r}
#7. I used the rbind() function to combine the four EPA datasets and the nrow() function to double check that there are 37,893 records.
EPA.bind <- rbind(EPAair_O3_NC2018_processed.csv, EPAair_O3_NC2019_processed.csv, EPAair_PM25_NC2018_processed.csv, EPAair_PM25_NC2019_processed.csv)

nrow(EPA.bind)

#8. I used a pipe function to wrangle this new dataset so that it fills the above-listed conditions using the filter(), group_by(), filter(), summarise(), and mutate() functions.  I then used the dim() function to double check that the dimensions of the dataset are 14,752 x 9.
EPA.bind.processed <-
  EPA.bind %>%
  group_by(Date, Site.Name, AQS_PARAMETER_DESC, COUNTY) %>%
  filter(Site.Name == "Linville Falls" | Site.Name == "Durham Armory" | Site.Name == "Leggett" | Site.Name == "Hattie Avenue" | Site.Name == "Clemmons Middle" | Site.Name == "Mendenhall School" | Site.Name == "Frying Pan Mountain" | Site.Name == "West Johnston Co." | Site.Name == "Garinger High School" | Site.Name == "Castle Hayne" | Site.Name == "Pitt Agri. Center" | Site.Name == "Bryson City" | Site.Name == "Millbrook School") %>%
  summarise(meanAQI = mean(DAILY_AQI_VALUE),
            meanLAT = mean(SITE_LATITUDE),
            meanLONG = mean(SITE_LONGITUDE)) %>%
  mutate(Month = month(Date),
         Year=year(Date))

dim(EPA.bind.processed)

#9. I used the pivot_wider() function to spread the datasets such that AQI values for ozone and PM2.5 are in separate columns.
EPA.bind.spread <- pivot_wider(EPA.bind.processed, names_from = AQS_PARAMETER_DESC, values_from = meanAQI)

#10. I used the dim() function to call up the dimensions of the new tidy dataset.  It is 8,976 x 9.
dim(EPA.bind.spread)

#11. I used the write.csv() function to save the processed dataset with the following file name: "EPAair_O3_PM25_NC1718_Processed.csv".
EPAair_O3_PM25_NC1718_Processed.csv <- EPA.bind.spread
write.csv(EPAair_O3_PM25_NC1718_Processed.csv, row.names = FALSE, file = "./Data/Processed/EPAair_O3_PM25_NC1718_Processed.csv")
```


## Generate summary tables
12. Use the split-apply-combine strategy to generate a summary data frame. Data should be grouped by site, month, and year. Generate the mean AQI values for ozone and PM2.5 for each group. Then, add a pipe to remove instances where a month and year combination (i.e. row) does not have corresponding ozone and PM2.5 data (use the function `drop_na` in your pipe). 

13. Call up the dimensions of the summary dataset.

```{r}
#12a. I used the split-apply-combine strategy to generate a summary data frame with the data grouped by site, month, and year.  I also used the summarise() function to generate the AQI values for ozone and PM2.5 for each group.
EPAair_O3_PM25_NC1718_Summary <-
  EPAair_O3_PM25_NC1718_Processed.csv %>%
  group_by(Site.Name, Month, Year) %>%
  summarise(meanOzoneAQI = mean(Ozone),
            meanPM2.5AQI = mean(PM2.5))

#12b. I added a pipe with the drop_na() function to remove instances where a month and year combination (i.e. row) does not have corresponding ozone and PM2.5 data.
EPAair_O3_PM25_NC1718_Summary <-
  EPAair_O3_PM25_NC1718_Processed.csv %>%
  group_by(Site.Name, Month, Year) %>%
  summarise(meanOzoneAQI = mean(Ozone),
            meanPM2.5AQI = mean(PM2.5))
  drop_na(EPAair_O3_PM25_NC1718_Summary)

View(EPAair_O3_PM25_NC1718_Summary)

#13. I used the dim() function to call up the dimensions of the summary dataset.  They are 308 x 5.
dim(EPAair_O3_PM25_NC1718_Summary)
```

14. Why did we use the function `drop_na` rather than `na.omit`?

> Answer: We used the function `drop_na` rather than `na.omit` because `drop_na` removes rows with missing data in any columns present.  The 'na.omit' function removes all incomplete cases of a singledata object.  In this case, we want to remove instances where a month and year combination does not have corresponding ozone and PM2.5 data.
